---
title: "DADA2_Ejemplo_Clase"
author: "Roberto Álvarez"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document: 
      fig_height: 8
      fig_width: 13
      number_section: yes
      theme: cerulean
      toc: yes
      toc_float:
        collapsed: no
        smooth_scroll: yes
---

# Datos

## Cargar los datos

```{r}
ibd_taxa <- read.csv('ibd_taxa.csv', row.names=1) # matriz de taxon x sample matrix, it means the percentage of the sample that corresponds to that taxa
ibd_lineages <- read.csv('ibd_lineages.csv', row.names=1) # lineage x taxon matrix, has the lineage of each sample
ibd_metadata <- read.csv('ibd_metadata.csv', row.names=1) # matriz de datos de lso metadatos x sample, metadata of the experiment 
```


## Ver los datos

```{r}
#the kable function makes a simple table, why look at them in this format??
knitr::kable(head(ibd_taxa, 10))
knitr::kable(head(ibd_lineages, 10))
knitr::kable(head(ibd_metadata, 10))
```


### Con tablas interactivas (Opcional)


```{r}
library(DT)
#didn't work, check later for the interactive tables
datatable(ibd_taxa)
datatable(ibd_lineages)
datatable(ibd_metadata)
```

## Barplot

```{r}
#this barplot shows the representation of each organism in the sabmple, a visual representation of the ibd_taxa
barplot(as.matrix(ibd_taxa))
```



```{r}
# generate number of colours equal to number of phyla
colours <- rainbow(length(unique(ibd_lineages$Phylum)))
color.indices <- match(ibd_lineages$Phylum, unique(ibd_lineages$Phylum))
# generate a vector with color for each taxon
colvec <- colours[color.indices]
```


This part will make the visual representation but with the colours that were assigned in the part right above. Must run all the chunk at the same time for it to work normally.
```{r}
barplot(as.matrix(ibd_taxa), col=colvec)
legend('topright', fill=colours,legend=unique(ibd_lineages$Phylum))
par(xpd=TRUE, mfrow = c(2,1), mar=c(1, 1, 1, 1))
plot(NULL ,xaxt='n',yaxt='n',bty='n',ylab='',xlab='', xlim=0:1, ylim=0:1)
legend("topright", fill=colours,legend=unique(ibd_lineages$Phylum), cex=.8)
barplot(as.matrix(ibd_taxa), col=colvec,  xaxt='n', yaxt='n', ylim=c(0, 1))
```


```{r}
library(ggplot2)
library(reshape2)
long_data <- cbind(ibd_taxa, ibd_lineages)#combine both matrices
long_data <- reshape2::melt(long_data)
```
```{r}
ggplot(data=long_data, aes(x=variable, y=value, fill=Phylum))+
  geom_bar(position='stack', stat='identity') + 
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) + labs(x='Sample', y='Relative abundance')
```



## Riqueza


```{r}
A <- rmultinom(5, size=100, prob=c(0.2, 0.4, 0.05, 0.02, 0.15, 0.13, 0.01, 0.04))
B <- rmultinom(5, size=100, prob=c(0.6, 0.25, 0, 0.04, 0.02, 0.06, 0.02, 0))


```


```{r}
counts <- cbind(A, B)
```


```{r}
groups <- c(rep('A', 5), rep('B', 5))
```



```{r}
toy_richness <- colSums(counts != 0)
```



```{r}
groups <- c(rep('A', 5), rep('B', 5))
wilcox.test(toy_richness~groups) 
```

```{r}
singletons <- colSums(counts == 1)
doubletons <- colSums(counts == 2)
rares <- singletons / (2*doubletons)
rares[doubletons == 0] <- 0

chao1 <- toy_richness + rares
```




```{r}
long_data <- reshape2::melt(ibd_metadata)
long_data = long_data[long_data$variable %in% c('shannon', 'pielou', 'simpson', 'invsimpson'),]
```




```{r}
ggplot(long_data, aes(x=Diagnosis, y=value, colour=variable)) + geom_boxplot()+
  facet_grid(cols=long_data$variable, scales='free')
```


```{r}
rarefaction <- function(count.vector=c(),depth=100){
probab=count.vector/sum(count.vector)
return(rmultinom(1,size=depth,prob=probab))
}
# we rarefy an example count vector
count.vector=rmultinom(n=1,size=150,prob=c(0.5,rep(0.1,5)))
rarefaction(count.vector)
```


## Medidas de diversidad

```{r}
sums <- apply(counts, 2, sum) # get the column sums (sample sums)
norm_counts <- counts # generate new data frame for normalized data
for (i in 1:ncol(counts)){ # divide per sample by sample total
  norm_counts [,i] <- counts[,i]/sums[i]
}
```



## Shannon

```{r}
shannon_div <- function(vector){
    vector <- vector*log(vector)
    # the vector has NA values if the species proportions are 0
    vectorsum <- sum(na.omit(vector))
    return((-1)*vectorsum)
}
```


```{r}
shannon_diversities <- apply(norm_counts, 2, shannon_div)
```

```{r}
ibd_metadata$shannon <- apply(ibd_taxa, 2, shannon_div)
```

## Visualizar las diferencias

```{r}
boxplot(shannon~Diagnosis, data=ibd_metadata, ylab='Shannon Diversity')
```



```{r}
long_data <- reshape2::melt(ibd_metadata)
long_data = long_data[long_data$variable %in% c('shannon', 'pielou', 'simpson', 'invsimpson'),]
```





```{r}
plot(shannon~Age, data=ibd_metadata, ylab='Shannon')
```



```{r}
cor(ibd_metadata$shannon, ibd_metadata$Age, method=c("spearman"))
cor.test(ibd_metadata$shannon, ibd_metadata$Age, method=c("spearman"))
```


## Comparar distribuciones de especies

```{r}
 fp.index=which(rownames(ibd_taxa)=="Faecalibacterium_prausnitzii")
wilcox.test(t(ibd_taxa[fp.index,ibd_metadata$Diagnosis=="Control"]),t(ibd_taxa[fp.index,(ibd_metadata$Diagnosis=="CD" || ibd_metadata$Diagnosis=="UC")]))

```

```{r}
col1=rgb(0,1,0,0.5)
col2=rgb(1,0,0,0.5)
# determine maximum density
maxy=max(hist(t(ibd_taxa[fp.index,]),breaks="FD",plot=FALSE)$density)
# draw first histogram for control samples
hist(t(ibd_taxa[fp.index,ibd_metadata$Diagnosis=="Control"]),breaks="FD", xlim=c(0,max(ibd_taxa[fp.index,])), ylim=c(0,maxy), prob=TRUE,col=col1, border=col1,xlab="Abundance", main="Histogram")
# draw second histogram for IBD samples
hist(t(ibd_taxa[fp.index,(ibd_metadata$Diagnosis=="CD" || ibd_metadata$Diagnosis=="UC")]),breaks="FD",prob=TRUE,col=col2, border=col2,add=TRUE)
# add a legend
legend("right",legend=c("Control","IBD"), lty = rep(1,2), col = c(col1,col2), merge = TRUE, bg = "white", text.col="black") 
```

## Medidas de disimilaridad

```{r}
library(vegan)
bray <- vegdist(t(norm_counts))
```


```{r}
pcoa.res.sim <- capscale(t(norm_counts)~1, distance='bray', na.action='na.omit')
eigenvalues <- eigenvals(pcoa.res.sim)
```



```{r}
plot(pcoa.res.sim)
```


```{r}
plot(pcoa.res.sim$CA$u[,c(1,2)],
     xlab=paste(c('PCoA1', round(eigenvalues[[1]],2))), ylab=paste(c('PCoA2', round(eigenvalues[[2]],2))))
```


```{r}
nmda.res.sim <- metaMDS(t(norm_counts), distance="bray", k=2, trymax=50)
stressplot(nmda.res.sim)


```



```{r}
plot(nmda.res.sim) 
```


```{r}
colvec = c(rep("#FF0000FF", 5), rep("#00FFFFFF", 5))
ordiplot(nmda.res.sim,type="n")
orditorp(nmda.res.sim, display="sites", col=colvec)
```


```{r}
points(nmda.res.sim, display="species", col='green')
```





```{r}
pcoa.res.ibd <- capscale(t(ibd_taxa)~1, distance='bray', na.action='na.omit')
eigenvalues <- eigenvals(pcoa.res.ibd)
plot(pcoa.res.ibd$CA$u[,c(1,2)],
     xlab=paste(c('PCoA1', round(eigenvalues[[1]],2))), ylab=paste(c('PCoA2', round(eigenvalues[[2]],2))))
```
```{r}
colvec = rainbow(length(unique(ibd_metadata$Diagnosis)))[as.numeric(as.factor(ibd_metadata$Diagnosis))]
plot(pcoa.res.ibd$CA$u[,c(1,2)], col=colvec,
     xlab=paste(c('PCoA1', round(eigenvalues[[1]],2))), ylab=paste(c('PCoA2', round(eigenvalues[[2]],2))),
     ylim=c(-0.25, 0.35))
legend("topleft", legend=unique(ibd_metadata$Diagnosis), col=unique(colvec), cex=1, pch=1)
```



```{r}
col_pal = colorRampPalette(c('red','blue'))(length(unique(ibd_metadata$Age)))
colvec = col_pal[ibd_metadata$Age]
plot(pcoa.res.ibd$CA$u[,c(1,2)], col=colvec,
     xlab=paste(c('PCoA1', round(eigenvalues[[1]],2))), ylab=paste(c('PCoA2', round(eigenvalues[[2]],2))))
```














```{r}
colvec = rainbow(length(unique(ibd_metadata$Diagnosis)))[as.numeric(ibd_metadata$Diagnosis)]
nmda.res.ibd <- metaMDS(t(ibd_taxa), distance="bray", k=2, trymax=50)
ordiplot(nmda.res.ibd,type="n")
points(nmda.res.ibd, display="sites", col=colvec)
```




